### 2020-08-20 thesis/thesis.tex
[thesis] Correct outdated sections

    \noindent
    TODO subset of this section's content is explained already

    These implications are straightforward when looking from a categorical standpoint,
    where effects are viewed as free models of algebraic theories~\cite{adequacy},
    and handlers are homomorphisms preserving the model structure~\cite{handlers}.
    Nevertheless, the results are very exciting for programming use cases. The current
    Freak implementation does not support I/O.

@@

    Examples for cohandlers

@@

    % Outdated
    As of this day, two implementations are available, one based on the curried
    translation and Appel~\cite{appel-continuations}, and the second one based
    directly on the uncurried translation with continuations as explicit stacks
    from paper. More details can be found in Section~\ref{chapter:implementation}.
    All commands are available within the \verb!src! directory.

@@

    After compiling and linking program to PATH, one may evaluate program as
    follows: \verb!freak programs/choicesList.fk!. The actual code is described in Section~\ref{sec:choice-example}

@@

    \noindent
    TODO Extend language with basic constructs for usability

    The syntax for the calculus is shown below. $nat \; n$ represents an integer $n$,
    `string' a string value, $V \oplus W$ and $V \approx W$ are respectively
    binary and relational operators, where we support basic arithmetic and
    comparison operations. \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$
    is a standard branching statement. The other constructs are just as in Links,
    with slight syntax modifications. Actual programs in Freak can be found in
    Section~\ref{sec:examples}.

@@

\section{Typing Rules}
\section{Operational Semantics}

    The source language's dynamics have been described
    extensively by providing small-step operational semantics,
    continuation passing style transformation~\cite{handlers-cps} as well
    as abstract machine~\cite{liberating-effects}, which was proved to coincide
    with CPS translation. That being said, Freak introduces new basic
    constructs to the language, for which we shall define the semantics.

@@

\section{Continuation Passing Style Transformation}
    \ldots

@@

    The Freak implementation is available on github~\cite{freak}, written purely
    in Haskell. Two inherently different takes at implementations were made. The
    first one is based on curried translation from A. Appel~\cite{appel-continuations}
    book, and the second one on Links language~\cite{handlers-cps, handlers-cps-journal},
    on the uncurried translation to target calculus with continuations represented
    as explicit stacks. We start by presenting core data structures, and
    afterwards move to actual translation details. Development of the former
    version is now suspended.

@@

    \noindent
    Evaluation of target calculus is typical to call-by-value with syntactic
    distinction between values and computations. For more details, refer
    to~\cite{handlers-cps-journal}. That being said, there is one part
    that requires more attention, namely, \verb!UTopLevelEffect Label UValue!,
    which represents an unhandled algebraic effect along with it's parameter.
    Even in case language has coalgebraic effects, special treatment is required
    for doing IO\@. For that reason, we define default handling for a few effects
    for printing to console and handling files, where semantic is just as one
    would expect from naming:

@@

\begin{verbatim}
    data ContComp
        = CPSApp CValue [CValue]
        | CPSResume CValue ContComp
        | CPSFix Var [Var] ContComp ContComp
        | CPSBinOp BinaryOp CValue CValue Var ContComp
        | CPSValue CValue
        | CPSLet Var CValue ContComp
        | CPSSplit Label Var Var CValue ContComp
        | CPSCase CValue Label Var ContComp Var ContComp
        | CPSIf CValue ContComp ContComp
        | CPSAbsurd CValue
\end{verbatim}
    Most of the terms at the end have a coinductive reference to itself, which
    represents the rest of the computation that needs to be done. For more
    clarification, one may take a look into the book mentioned
    above~\cite{appel-continuations}. The source code for curried translation and
    evaluation can be found respectively in \verb!CPSAppel.hs! and
    \verb!EvalCPS.hs!.

@@

    Having in mind the drawbacks mentioned above, alternative translation was
    written, that coincides with the translation from the paper. Namely, with
    the uncurried translation to target calculus with continuations represented
    as explicit stacks. The target calculus was described in Section~\ref{sec:implementation-ast},
    for which the evaluation is straightforward. The continuations are represented
    as \verb!Cont!, with syntactic distinction between pure and effectful computations,
    which occupy alternating positions in the stack. Explicit distinction gave
    more control in the source code.

@@

    data Cont = Pure ContF
              | Eff ContF
              | Coeff ContF

    initialPureCont :: ContF
    initialPureCont v ks = (return . UVal) v

    initialEffCont :: ContF
    initialEffCont (UPair (ULabel effLabel) (UPair p r)) ks =
        return . UApp (UVal r) $ UTopLevelEffect effLabel p
\end{verbatim}
    Where \verb!CPSMonad! is a monad transformer over \verb!Either!, \verb!State!
    and \verb!IO!. \verb!State! is required to generate labels for fresh variables
    that came from the translation, \verb!Either! for handling exceptions that
    may occur during translations, and \verb!IO! for handling input output of the
    source language. Initial pure continuation is lifting values into computations,
    and effectful one is propagating previously mentioned \verb!UTopLevelEffect!.
    The core code is split into a few functions:

@@

    \noindent
    Where the first two are implementing cps for computations and values.
    \verb!cpsOp! is doing translation for an (co) effect that occured in program,
    \verb!cpsHRet! and \verb!cpsHOps! are yielding pure and effectful continuations,
    based on a given handler. \verb!forward! is responsible for forwarding the
    computation to the outer handler, and the last one for running the
    translation, where evaluation result is represented by\\
    \verb!type EvalResMonad a = IO (Either Error a)!. This results in an
    implementation that finally supports nested handlers, which can be seen by
    running tests.

>>>>

    These implications are straightforward when looking from a categorical standpoint,
    where effects are viewed as free models of algebraic theories~\cite{adequacy},
    and handlers are homomorphisms preserving the model structure~\cite{handlers}.
    Nevertheless, the results are very exciting for programming use cases.

@@

    TODO Examples for cohandlers

@@

    % Outdated
    As of this day, two implementations are available, one based on the curried
    translation from Appel~\cite{appel-continuations}, and the second one based
    directly on the uncurried translation with continuations as explicit
    stacks~\cite{handlers-cps-journal}. More details can be found in
    Section~\ref{chapter:implementation}. All commands are available within the
    \verb!src! directory.

@@

    After compiling and linking program to PATH, one may evaluate program as
    follows: \verb!freak programs/choicesList.fk! The actual code is described in Section~\ref{sec:choice-example}

@@

    The syntax for the calculus is shown below. $nat \; n$ represents an integer $n$,
    `string' a string value, $V \oplus W$ and $V \approx W$ are respectively
    binary and relational operators, where we support basic arithmetic and
    comparison operations. \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$
    is a standard branching statement. The other constructs are just as in
    Links~\cite{handlers-cps-journal}, with slight syntax modifications.
    Actual programs in Freak can be found in Section~\ref{sec:examples}.

    TODO cohandlers

@@

\section{Dynamics}

    Semantic of Freak is heavily based on Links langauge~\cite{handlers-cps-journal},
    for which the source language's dynamics have been described extensively by
    providing small-step operational semantics, continuation passing style
    transformation~\cite{handlers-cps} as well as abstract
    machine~\cite{liberating-effects}, which was proved to coincide with CPS
    translation. That being said, Freak introduces new basic constructs to the
    language, for which we shall define the semantics.

@@

    TODO cohandlers

@@

    The Freak implementation is available on github~\cite{freak}, written purely
    in Haskell. Two inherently different takes at implementations were made. The
    first one is based on curried translation from A. Appel~\cite{appel-continuations}
    book, and the second one on Links language~\cite{handlers-cps, handlers-cps-journal},
    on the uncurried translation to target calculus with continuations represented
    as explicit stacks. We start by presenting core data structures, and
    afterwards move to actual translation details. Development of the former
    Appel's version is now suspended.

@@

    \noindent
    Evaluation of target calculus is typical to call-by-value with syntactic
    distinction between values and computations. For more details, refer
    to~\cite{handlers-cps-journal}. That being said, there is one part
    that requires more attention, namely, \verb!UTopLevelEffect Label UValue!,
    which represents an unhandled algebraic effect along with it's parameter.
    Even in case language has coalgebraic effects, special treatment is required
    for doing IO\@. For that reason, we define default handling for a few effects
    for printing to console and handling files, where semantic is just as one
    would expect from naming. We abuse haskell's notation to define types for
    algebraic effects living in Freak langauge:

@@

\begin{verbatim}
    data ContComp
        = CPSApp CValue [CValue]
        | CPSResume CValue ContComp
        | CPSFix Var [Var] ContComp ContComp
        | CPSBinOp BinaryOp CValue CValue Var ContComp
        | CPSValue CValue
        | CPSLet Var CValue ContComp
        | CPSSplit Label Var Var CValue ContComp
        | CPSCase CValue Label Var ContComp Var ContComp
        | CPSIf CValue ContComp ContComp
        | CPSAbsurd CValue
\end{verbatim}
    Most of the terms at the end have a coinductive reference to itself, which
    represents the rest of the computation that needs to be done. For more
    clarification, one may take a look into the book mentioned
    above~\cite{appel-continuations}. The source code for curried translation and
    evaluation can be found respectively in \verb!CPSAppel.hs! and
    \verb!EvalCPS.hs!. Development of this version of translation is discontinued.

@@

    Having in mind the drawbacks mentioned above, alternative translation was
    written, that coincides with the translation from~\cite{handlers-cps-journal}.
    Namely, with the uncurried translation to target calculus with continuations
    represented as explicit stacks. The target calculus was described in
    Section~\ref{sec:implementation-ast}. The continuations are represented as
    \verb!Cont!, with syntactic distinction between pure, effectful and
    coeffectful computations. Pure and (co) effectful continuations occupy
    alternating positions in the stack. Explicit distinction was made to provide
    more control in the source code.

@@

    data Cont = Pure ContF
              | Eff ContF
              | Coeff ContF
\end{verbatim}

    \noindent
    Where \verb!CPSMonad! is a monad transformer over \verb!Either!, \verb!State!
    and \verb!IO!. \verb!State! is required to generate labels for fresh variables
    that came from the translation, \verb!Either! for handling exceptions that
    may occur during translations, and \verb!IO! for handling input output of the
    source language.

\begin{verbatim}
    initialPureCont :: ContF
    initialPureCont v ks = (return . UVal) v

    initialEffCont :: ContF
    initialEffCont (UPair (ULabel effLabel) (UPair p r)) ks =
        return . UApp (UVal r) $ UTopLevelEffect effLabel p

    initialContStack :: [Cont]
    initialContStack = [Pure initialPureCont, Eff initialEffCont]
\end{verbatim}

    \noindent
    Initial pure continuation is lifting values into computations,
    and effectful one is propagating previously mentioned \verb!UTopLevelEffect!.
    Notice that \verb!initialContStack! does not have initial cohandler,
    as handling of default coeffects is embed in evaluator. The core code is
    split into a few functions:

@@

    \noindent
    Where the first two are implementing cps for computations and values.
    \verb!cpsOp! is doing translation for an (co) effect that occured in program,
    \verb!cpsHRet! and \verb!cpsHOps! are yielding pure and effectful continuations,
    based on a given handler. \verb!forward! is responsible for forwarding the
    computation to the outer handler, and the last one for running the
    translation, where evaluation result is represented by\\
    \verb!type EvalResMonad a = IO (Either Error a)!. This results in an
    implementation that finally supports nested handlers, which can be seen by
    running tests. Links language~\cite{handlers-cps-journal} also supports
    shallow handlers~\cite{shallow-handlers}, whereas our langauge implements
    only deep ones.
